Endogenous long-run output impacts of monetary and fiscal policy: a growth theory approach

### 2021-12-07 paper.tex
simplified analysis

\subsection{Definitions}
There are three homogenous pools of labor.  \(L_1(t)\) is labor that, combined with capital, produces output.  \(L_2(t)\) is labor that works in research and development.  This labor creates technological efficiencies, \(A(t)\), which enhance the productivity of \(L_1(t)\).  Finally, there is \(L_3(t)\), which is either the ``bureaucratic'' labor which maintains the administrative state, or represents capable labor that chooses not to participate in the labor force.  Since \(L_3(t)\) does not impact the output, the model is agnostic to whether labor force participation or bureaucratic administration is the primary driver of \(L_3(t)\).  There is capital \(K(t)\).  \(i\) is the interest rate set by the central bank, and \(g\) is the fraction of output that is used for government spending.  It is assumed that this spending is neutral from an output perspective: that is, labor and capital that is employed through fiscal policy has the same output as private sector labor and capital and in the same proportions.
\\
\\
Following Solow \cite{solow}, output \(Y(t)\) takes the following form:

@@

Here \(\gamma_0\) is the non-intervention (baseline) rate of growth in research labor, \(\gamma_1\) is the sensitivity of labor to the prevailing interest rate's deviation from the ``natural'' rate \(i^*\), and \(\gamma_2\) is the proportion of government spending that is dedicated to research and development.  For instance, federal grants to universities would be considered spending on basic research.   
\\
\\
\(L_1\) is determined by the above equations:
\[\frac{dL_1(t)}{dt}=\frac{dL}{dt}-\frac{dL_2(t)}{dt}-\frac{dL_3(t)}{dt}\] 
\[=\eta\left(L_1(t)+L_2(t)+L_3(t)\right)-\frac{dL_2(t)}{dt}-\frac{dL_3(t)}{dt}\]
\\
\\
Technology dynamics are directly proportional to the labor dedicated to research.  While in reality research requires some level of capital, the assumption simplifies the analysis.  Many kinds of research (for example, mathematics and economics) requires minimal capital investment.  Of course, the physical sciences often require massive capital investments.  This is a shortcoming of the model.  Note that this also deviates from the classic Solow model where technological growth is exogoneous.
\[\frac{dA(t)}{dt}=\beta L_2(t)\]

@@

\subsubsection{Economic collapse}
If either \(L_2\) or \(L_3\) grow too large, either \(K\) or \(L_1\) can shrink to zero.  This can occur if interest rates are held too low or \(g\) is too large.  \(L_2\) will grow too quickly if \(\gamma_0+\gamma_1 (i^*-i)+\gamma_2 g > \eta\).  Likewise, \(L_3\) will crowd out \(L_1\) if \(v_0+v_1g > \eta\) and \(g\) IS SOMETHING.

@@

Since the model is a system of non-linear differential equations, the model must be solved numerically.  I show plots of two economies under different interest rate and fiscal policy regimes.  The full parameters for these two economies:

\begin{center}
\begin{tabular}{ c c c }
      Variable  & Economy 1 Value & Economy 2 Value \\ 
      \(\eta\) & 0.02 & 0.02 \\  
      \(\gamma_0\) & 0.004 & 0.04  \\  
      \(\gamma_1\) & 0.2 & 0.2  \\  
      \(\gamma_2\) & 0.005 & 0.005  \\ 
      \(v_0\) & 0.001 & 0.001  \\  
      \(v_1\) & 0.05 & 0.05  \\  
      \(\beta\) & 0.03 & 0.03  \\  
      \(\alpha\) & 0.4 & 0.4  \\  
      \(\delta\) & 0.1 & 0.1  \\  
      \(s\) & 0.3 & 0.3  \\ 
      \(L_1(0)\) & 3.0 & 3.0  \\ 
      \(L_2(0)\) & 1.0 & 1.0  \\ 
      \(L_3(0)\) & 1.0 & 1.0  \\ 
      \(A(0)\) & 2.0 & 2.0  \\ 
      \(K(0)\) & 10.0 & 10.0  
\end{tabular}
\end{center}

\subsection{Low base growth of research}

In the first economy, the base growth rate of \(L_2\) is low, providing quite a bit of freedom for monetary policy.

@@

\begin{minipage}{\linewidth}
\begin{framed}
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_0_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_1_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_2_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_3_labor}
\end{minipage} \hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_4_labor}
\end{minipage}\hfill
\end{framed}
\end{minipage}

@@

\subsection{High base growth of research}

In the second economy, the base growth rate of \(L_2\) is high, so that even mild interest rate regimes have deleterious long run impacts.

\begin{minipage}{0.8\linewidth}
\begin{framed}
\centering
\includegraphics[width=1\textwidth]{images/economy_1}
\end{framed}
\end{minipage}

\begin{minipage}{\linewidth}
\begin{framed}
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_1_run_0_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_1_run_1_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_1_run_2_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_1_run_3_labor}
\end{minipage} \hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_1_run_4_labor}
\end{minipage}\hfill
\end{framed}
\end{minipage}

>>>>

\subsection{Definitions}
There are three homogenous pools of labor.  \(L_1(t)\) is labor that, combined with capital, produces output.  \(L_2(t)\) is labor that works in research and development.  This labor creates technological efficiencies, \(A(t)\), which enhance the productivity of \(L_1(t)\).  Finally, there is \(L_3(t)\), which is either the ``bureaucratic'' labor which maintains the administrative state, or represents capable labor that chooses not to participate in the labor force.  Since \(L_3(t)\) does not impact the output, the model is agnostic to whether labor force participation or bureaucratic administration is the primary driver of \(L_3(t)\).  There is capital \(K(t)\).  \(i\) is the interest rate set by the central bank, which is assumed to be exogenously shocked and then reverts to the natural rate.  \(g\) is the fraction of output that is used for government spending.  It is assumed that this spending is neutral from an output perspective: that is, labor and capital that is employed through fiscal policy has the same output as private sector labor and capital and in the same proportions.
\\
\\
Following Solow \cite{solow}, output \(Y(t)\) takes the following form:

@@

Here \(\gamma_0\) is the non-intervention (baseline) rate of growth in research labor, \(\gamma_1\) is the sensitivity of labor to the prevailing interest rate's deviation from the ``natural'' rate \(i^*\), and \(\gamma_2\) is the proportion of government spending that is dedicated to research and development.  For instance, federal grants to universities would be considered spending on basic research.  \(i\) has the following dynamics:

\[\frac{di(t)}{dt}=a\left(i^*-i\right)\]
\\
\\
\(L_1\) is determined by the above equations:
\[\frac{dL_1(t)}{dt}=\frac{dL}{dt}-\frac{dL_2(t)}{dt}-\frac{dL_3(t)}{dt}\] 
\[=\eta\left(L_1(t)+L_2(t)+L_3(t)\right)-\frac{dL_2(t)}{dt}-\frac{dL_3(t)}{dt}\]
\\
\\
Technology dynamics are directly proportional to the labor dedicated to research.  While in reality research requires some level of capital, the assumption simplifies the analysis.  Many kinds of research (for example, mathematics and economics) requires minimal capital investment.  Of course, the physical sciences often require massive capital investments.  This is a shortcoming of the model.  Note that this also deviates from the classic Solow model where technological growth is exogoneous.
\[\frac{dA(t)}{dt}=\beta L_2(t)\]

@@

\subsubsection{Economic collapse}
If either \(L_2\) or \(L_3\) grow too large, either \(K\) or \(L_1\) can shrink to zero.  This can occur if interest rates are shocked too low or \(g\) is too large.  \(L_2\) will grow too quickly if \(\gamma_0+\gamma_1 (i^*-i)+\gamma_2 g > \eta\).  Likewise, \(L_3\) can crowd out \(L_1\) if \(v_0+v_1g > \eta\).

@@

Since the model is a system of non-linear differential equations, the model must be solved numerically.  I show plots of an economy under different interest rate and fiscal policy regimes.  The full parameters for this economy:

\begin{center}
\begin{tabular}{ c c }
      Variable  & Value  \\ 
      \(\eta\) & 0.02 \\  
      \(\gamma_0\) & 0.004   \\  
      \(\gamma_1\) & 1.5   \\  
      \(\gamma_2\) & 0.005   \\ 
      \(v_0\) & 0.001  \\  
      \(v_1\) & 0.05  \\  
      \(a\) & 0.1  \\
      \(\beta\) & 0.03   \\  
      \(\alpha\) & 0.4   \\  
      \(\delta\) & 0.1   \\  
      \(s\) & 0.3   \\ 
      \(L_1(0)\) & 3.0   \\ 
      \(L_2(0)\) & 1.0  \\ 
      \(L_3(0)\) & 1.0   \\ 
      \(A(0)\) & 2.0  \\ 
      \(K(0)\) & 10.0   
\end{tabular}
\end{center}

Monetary policy in either direction can lead to low growth or collapse.  An increase in interest rates can lead to lower levels of technology improvement, which can slow economic growth.  A decrease in interest rates can lead to unsustainable growth in technology labor, leading to economic collapse.  Similarly, as fiscal policy gets larger, more of the productive economy is ``crowded out''.  At very high levels of government spending, the economy tends towards collapse as non-productive labor increases.

@@

\begin{minipage}{\linewidth}
\begin{framed}
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_0_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_1_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_2_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_3_labor}
\end{minipage} \hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_4_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_5_labor}
\end{minipage}\hfill
\begin{minipage}[t]{.48\textwidth}
\centering
\includegraphics[width=1\textwidth]{images/econ_0_run_6_labor}
\end{minipage}\hfill
\end{framed}
\end{minipage}

@@
